Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-10-01 papers/paper1/case_study.tex
Updates through page 10.

In this section, we illustrate how ioa++ can be used to build real distributed systems.
In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
The lower layers provide services such as TCP/UDP sockets (for communicating between endsystems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
The intermediate layers, i.e., frameworks like ioa++, contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.

>>>>

In this section, we illustrate how ioa++ can be used to build real distributed systems.
In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
The intermediate layers, i.e., frameworks like ioa++, contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.

### 2011-10-01 papers/paper1/programming_model.tex
Updates through page 10.

A number of wrappers and macros are provided by ioa++ for declaring actions.
%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
%%The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly except for the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata are declared public.

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
Another common idiom is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
\ifjournal
For the asynch\_lcr\_automaton, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this common member function.
\fi

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received if produced.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

@@

\paragraph*{System action events}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

>>>>

A number of wrappers and macros are provided by ioa++ for declaring actions.
%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in Section~\ref{practical} are available.
Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
%%The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
\ifjournal
For example, the send action of the asynch\_lcr\_automaton is declared:
\begin{lstlisting}
private:
  bool send_precondition () const { ... }
  UID send_effect () { ... }
  void send_schedule () const { ... }
public:
  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
\end{lstlisting}
\fi
Input actions are declared similarly except for the absence of a precondition.
The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
The macros for actions that are subject to binding by other automata are declared public.

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{scheduling}.

\ifjournal
For the asynch\_lcr\_automaton, the scheduling function is:
\begin{lstlisting}
  void schedule () const {
    if (send_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::send);
    }
    if (leader_precondition ()) {
      ioa::schedule (&asynch_lcr_automaton::leader);
    }
  }
\end{lstlisting}
The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this common member function.
\fi

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in Section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of output actions to ensure that the value generated by the output action will be received if produced.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

@@

\paragraph*{System action events}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in Section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.
